---
title: "EAB 770 - Chapter 1"
author: "Miguel Fudolig"
format:
  revealjs:
    theme: clean.scss
    scrollable: true
    footer: "Chapter 1 - [Back to home](https://madfudolig.quarto.pub/eab770slides/)"
    code-block-bg: true
    code-block-border-left: "#31BAE9"
    slide-number: true
    menu: true
    chalkboard: true
    engine: knitr
    echo: true
  pdf: 
    number-sections: true
---

# Chapter 1: Introduction

## Outline

- Syllabus Discussion
- Accessing `SAS` OnDemand for Academics
- Why are we here?
- Preliminary Materials

## Syllabus

- The syllabus can be accessed through
    - Canvas
    - <https://miguelfudolig-eab-770-syllabus.share.connect.posit.cloud>

## Programming Language

- We will primarily use R and `SAS` for this course.
    - You can access `SAS` through `SAS` On Demand for Academics. This will be free and accessible when you have internet connection.
    - You can download the most recent versions of `R` and `RStudio` if you prefer to take the course in `R`. You can access a quick tutorial on how to install these [here](https://rstudio-education.github.io/hopr/starting.html).

## Accessing `SAS` On Demand for Academics

- Create a `SAS` Profile that will be used to register for `SAS` OnDemand for Academics. To register, visit <https://www.sas.com> and fill out the form to *Create profile*.

  - Use your UNLV email for faster verification.

- Create your account for `SAS` OnDemand for Academics. To register, visit <https://welcome.oda.sas.com> sign in with your `SAS` Profile and complete the steps to *Register for an account*.


# Why are we here?

Why bother learning categorical data analysis?


# Categorical Data

## Categorical Data

Categorical data is present everywhere!

- Demographics
- Opinions
- Ratings
- Locations

## Types of Categorical Response Variables

- Dichotomous

  - Yes/No

- Ordinal

  - More than two possible outcomes with inherent ordering

- Nominal

  - More than two possible outcomes without inherent ordering

- Discrete Counts

  - Numerical outcomes consisting of counting numbers.

- Grouped Survival Times

  - Number of patients who fail/survive during a specific time interval.
  - The interval of failure/survival (time-to-event) is categorical.

## Sampling Frameworks

- Historical data
    - Observational data, usually secondary data
    - No randomization
    - Example: Public data sets.
- Experimental Data
    - Studies that involve random assignment of treatments to different subjects.
    - Random assignment/allocation
    - Example: Randomized Controlled Trials
- Sample Survey Studies
    - Randomly sampled from a larger study population
    - Random selection

## Effect of Randomization

- Randomization affects conclusions of generalization and causation.
- Historical data
    - No randomization: hard to generalize unless the data was randomly sampled from a predetermined population
- Experimental Data
    - Random assignment: balances effects of confounding variables
- Sample Survey Studies
    - Random sampling: representative sample provides good coverage of the larger population

## Randomization

Randomization can be done per subject or a cluster of subject.

- Strata (survey studies) or blocks (experiments)
- Examples
    - Stratified random samples
    - Randomized block designs
    - Multi-level experimental designs

## `SAS` Procedures and Data

- There are multiple ways to import data into `SAS`
    - DATA step
    - PROC IMPORT step
    - `SAS` OnDemand can import data from CSV files.
- `FREQ`
    - Contingency tables ($2\times 2$, $s \times r$)
    - Chi-squared tests
- `LOGISTIC`
    - Logistic Regression
- `GENMOD`
    - Generalized Estimating Equations
- Other procedures include `GLM` and `CATMOD`.

## `SAS` in Class

- Sample code will be provided, but there will be examples where we work the code out together as a class.
- Class computers have `SAS` installed in them.
    - You can save files on your Y drive (assigned to your ACE ID) to access files on other computers.
    - `SAS` can also be accessed through `SAS` OnDemand Academics.
        - Needs internet connection

## `SAS` Examples

- Download Chapter1.sas from Canvas.

## `SAS` Example

::: panel-tabset

### Coding Rules
- ALL lines should be ended with a semi-colon **;**
- `data` statement is followed by the data set name.
- `length` specifies the number of characters included in each **character**.
- `input` names all variables included in the data set
    - The `$` after the variable tells `SAS` that these variables are categorical variables.
- `datalines;` tells `SAS` that you are about to add new data points.
    - Each row represents a data point in this data table.
    - End with a separate **;**

### Example
```
data respire;
length treat $7. outcome $1. ;
input treat $ outcome $ count;
datalines;
placebo f 16
placebo u 48
test f 40
test u 20
;
```
:::




## `SAS` Example: Contingency Table

::: columns
::: {.column style="width:50%; font-size: 125%"}


```
proc freq data=respire;
weight count;
tables treat*outcome;
run;
```

:::

::: {.column style="width:50%; font-size: 75%"}
- `proc freq` calls the `FREQ` procedure in `SAS`.
  - Not defining `data=respire` means `SAS` will automatically use the most recently run data set.
- `weight` tells `SAS` the number of counts for each count
- `tables` creates the contingency table with row, column, and total percentages.
- `run;` tells `SAS` to run the code. **This is important!**
:::
:::

## `SAS` Example: One Observation per Row

::: columns
::: {.column style="width:50%; font-size: 100%"}


```
data severe;
input treat $ outcome $ @@;
datalines;
placebo f placebo f placebo f
placebo f placebo f
placebo u placebo u placebo u
placebo u placebo u placebo u
placebo u placebo u placebo u
placebo u
test f test f test f
test f test f test f
test f test f
test u test u test u
test u test u test u
test u test u test u
test u test u test u
test u test u test u
test u test u test u
test u test u
;
```

:::

::: {.column style="width:50%; font-size: 75%"}
- This is an example of a data set where one row corresponds to one observational unit.
- The `@@` sign means that the data lines contain multiple observations. `@@` basically means new line after two entries written in the `datalines` statement.
:::
:::

## `SAS` Example: Contingency Table

::: columns
::: {.column style="width:50%; font-size: 100%"}


```
proc freq data=severe;
tables treat*outcome / nocol nopct;
run;
```

:::

::: {.column style="width:50%; font-size: 75%"}
- The `weight` statement is not needed in the `FREQ` procedure when the data does not consist of aggregate counts.
- `nocol` suppresses the column percentages in the output
- `nopct` suppresses the total percentages in the output
- `norow` suppresses the row percentages in the output
:::
:::

# `R`

## Importing files

There are multiple ways to import files into `R`. The most common data set format is the `.csv` file. You can import CSV files using the `read.csv(filename)` function.

First, download the file `SleepHealthData.csv` from Canvas in the same folder as your R source code.

```{r}
#| execute: false
getwd() # save your data in this folder
adverse <- read.csv("adverse.csv") # or use a specific path to your folder
adverse
```

## Data Frames



Data frames are rectangular spreadsheets. Data are typically imported as data frames.

::: callout-note
Rows correspond to observations and the columns correspond to variables.
:::

Example:
```{r}
head(mtcars)

sample_data  <- data.frame(ID = c(1,2,3), Name = c("Alice","Bob","Charlie"))
sample_data
```

## `R` packages

`R` packages extend the functionality of `R` by providing additional functions, data, and documentation. These packages are written by `R` users around the world and can be downloaded for free!

::: callout-note
Think of `R` as a new phone. `R` packages are apps that you can download to use your phone in many different ways.
:::

## Installing and Loading `R` Packages

Like apps on a phone, `R` packages also need to be installed. These packages can be installed by running the following code snippet `install.packages("PackageName")`. For example, to install the package `tidyverse` used for data manipulation and cleaning, you can run the following code:

```{.r}
install.packages("tidyverse")
```

To load this package in `R`, you can use the following syntax:

```{r}
library(tidyverse)
```

::: callout-important
You must have an active internet connection to install `R` packages to your device.
:::

## Contingency Tables

To create your own contingency table, you can use the `data.frame()` and `xtabs()` functions.

```{r}
count_data <- data.frame(Row = c("A","A","B","B"),
                         Column = c("X","Y","X","Y"),
                         Count = c(2,4,6,8))

xtabs(Count~Row+Column,data=count_data)
```


##  Exercise


Suppose a survey asking sophomores and senior students whether they agree with the statement “I enjoyed my in-person classes this semester.” yielded the following results.

Enter the data in `SAS` using the data step and generate the contingency tables. Name the data set `enjoy`.

|           | Strongly Agree | Agree | Neutral | Disagree | Strongly Disagree |
|:---------:|:--------------:|:-----:|:-------:|:--------:|:-----------------:|
| Sophomore |       24       |  43   |   12    |    25    |        12         |
|  Seniors  |       19       |  37   |   24    |    32    |        21         |



::: panel-tabset
### `R` Code

```{r}
enjoy <- data.frame(year = c(rep(c("Sophomore","Seniors"),each=5)),
                    response = c(rep(c("SA","A","N","D","SD"),by=5)),
                    count = c(24,43,12,25,12,19,37,24,32,21))

xtabs(count~year+response,data=enjoy)
```


### `SAS` Code




There are many ways to code the `data` step. Here's one way to do it!

```
data enjoy;
length year $9. response $2.;
input year $ response $ count @@;
datalines;
sophomore SA 24 sophomore A 43
sophomore N 12
sophomore D 25 sophomore SD 12
senior SA 19 senior A 37
senior N 24
senior D 32 senior SD 21
;

proc freq order=data data=enjoy;
weight count;
tables year*response;
run;
```

:::


